### 2013-07-28 patchy-selection-paper.tex
.

Migration--selection balance ensures that 
there will be some number of $B$ alleles present in the regions where they are disadvantageous,
but only very few, and rarely, far away from the patch where $B$ is established.
Denote the expected frequency of allele $B$ at distance $r$ from the patch by $q(r)$.
Following \citet{slatkin1973geneflow}, one can write down a differential equation to which $q(r)$ is the solution,
which can then be solved to give $q(r)$ in terms of Jacobi elliptic functions (see Appendix \ref{apx:elliptic_integrals}).
Asymptotics of these implies that for large $r$ the mean occupation frequency decays exponentially, 
\begin{align} \label{eqn:eqfreq}
    q(r) \approx C \exp( -\sqrt{2 s_m} r / \sigma) \qquad \text{for large $r$},
\end{align}
where $C$ is a constant depending on the geographic shape of the populations and the selection coefficients. 
In Appendix \ref{apx:patchy_sims} we argue that $C=1/2$ is about right for most parameter values,
and also that these asymptotics are correct in both one and two dimensions
(the differential equation is only analytically solvable in one dimension).

@@

The arguments depends on the following decomposition of the genealogy of $B$ alleles.
First, fix some intermediate distance $r_0$ from the established patch,
and classify all $B$ alleles present at any time outside $r_0$ into families
if they share a common ancestor living outside $r_0$.
If $r_0$ is large enough that there are not too many families
and interactions between family members are weak,
then these families can be approximated by a collection of independent spatial branching processes.
(This can be made formal in the limit of large population density.)
We can extend these families by 
including any individual outside the original patch (but perhaps inside $r_0$) 
that has a parent outside $r_0$,
so that these branching processes are rooted at $r_0$, and are killed upon reaching the original patch.
This decomposition is illustrated in figure \ref{fig:sim_excursions}.
In terms of these families, the equilibrium frequency above is then
\begin{align}
    \label{eqn:gestalt_q}
    \begin{split}
        q(r) &= \text{ ( outflux of families ) } \times \text{ ( mean occupation density of a family at $r$ ) } ,
\end{split}
\end{align}
and the quantity we wish to compute,
the effective rate of at which \emph{families of} migrant $B$ alleles establish in a new patch at distance $r$, 
is
\begin{align}
    \label{eqn:gestalt_migrate}
    \begin{split}
        \migrate(r) &= \text{ ( outflux of families ) } \times \text{ ( probability a family establishes in patch at $r$ ) }
    \end{split}
\end{align}
All terms except the ``outflux of families'' are simple calculations with spatial branching processes;
and we can solve for this outflux using the established form for \eqref{eqn:gestalt_q}.
Here $\migrate$ reflects the mean number of establishing migrant families per unit time,
rather than individuals,
because family establishment events may include more than one individual,
so individual establishment events are clustered in time,
and using their overall density to obtain a typical waiting time until the next event
will underestimate that time.

@@

\begin{figure}[ht!!]
  \begin{center}
  \input{branching-concept.pdf_tex}
  \end{center}
\caption{Depiction of the decomposition described in the text.
In red is the central lineage (whose spatial motion we follow);
in black are the surviving branches off of this lineage, 
and in grey are offspring of the central lineage that did not survive until $t$.
\plr{Label $W_s$ in this figure?}
\plr{Add as part B a simulation figure.}
\label{fig:branching_decomp}
}
\end{figure}

@@

\begin{figure}[ht!]
    \begin{center}
        % made in patchy-sim-writeup-plots.R
        \includegraphics{example-equilibrium}
    \end{center}
    \label{fig:prob_estab_calcs}
    \caption{
    Snapshots of a simulation of allelic frequency dynamics around a patch where the allele is beneficial:
    the focal allele is benefical in the middle region and deleterious outside.
    The dynamics were simulated for 10000 generations across demes with 1000 individuals in each
    and random dispersal to nearby demes with a mean dispersal distance of $\sigma \approx 2$ demes.
    Colored lines show allele frequencies across space in 50 uniformly spaced generations;
    the solid line shows the mean across all generations.
    Below is the same plot, except on a log scale.
    }
\end{figure}

@@

One route to the ``equlibrium frequency'' of the allele outside the range where it is advantageous is as follows;
see \citet{slatkin1973geneflow} or \citet{barton1987establishment,pollak1966survival} for other arguments in this case, 
or \citet{etheridge} or \citet{dawson} for general theory used below.
Suppose that the population is composed of a finite number of small demes of equal size $N$ arranged in a regular grid,
and that selection (for or against) the allele is given by the function $s(x)$, with $x$ denoting the spatial location.
Then each individual at location $x$ reproduces after a random lifetime,
dying and giving birth in the same deme to a random number of offspring with distribution given by $X$;
who all migrate to a new location chosen as $x+R$
and replace randomly chosen individuals there.
If $x+R$ is outside of the range, then they perish.
Each individual's reproduction time is exponentially distributed, 
either with rate 1 if it carries the original allele, or with rate $1+s(x)$ if it carries the mutant allele.
Suppose that $R$ has mean zero and variance $\sigma^2$, and that $X$ has mean $\mu$ and variance $\xi^2$.
We also assume that the distribution of $X$ makes the resulting Markov chain irreducible and aperiodic.

Let $\Phi^N_t(x)$ be the number of mutant alleles present at location $x$ at time $t$,
and denote by $\delta_x$ a single unit at location $x$, so that e.g.~$\Phi^N_t + \delta_x/N$
is the configuration after a mutant allele has been added to location $x$.
For $0\le \phi \le 1$, we also denote by $\bar X_\phi$ the number of mutant alleles added if $X$ new offspring carrying mutant alleles
replace randomly chosen individuals in a deme where the mutant allele is at frequncy $\phi$ (i.e.~hypergeometric with parameters $(X,\phi)$);
similarly, $\tilde X_\phi$ is the number lost if the new offspring do not carry the allele (i.e.~hypergeometric with parameters $(X,1-\phi)$).
(We like to think of $\Phi^N_t$ as a measure, but it does not hurt to think of $\Phi^N$ as a vector;
we aren't providing the rigorous justification here.)
Then we know that for any sufficiently nice function $f(\phi)$ that
\begin{align} \label{eqn:discrete_generator}
  \begin{split} \frac{\partial}{\partial t} \E\left[ f(\Phi^N_t) \right] 
  &= N \sum_x \left\{ \E\left[ (1+s(x+R)) \Phi^N_t(x+R) \left( f\left(\Phi^N_t + \frac{\bar X_{\Phi_t(x+R)}}{N}\delta_{x}\right) - f(\Phi^N_t) \right) \right] \right. \\
     & \qquad  \qquad \left. {} + \E\left[ \left(1-\Phi^N_t(x+R)\right) \left( f\left(\Phi^N_t - \frac{\tilde X_{\Phi_t(x+R)}}{N}\delta_{x}\right) - f(\Phi^N_t) \right) \right] \right\}  \end{split} \\
     &= \mu \sum_x \E\left[ \left(\partial_{\phi(x)} f(\Phi_t) \right) \left\{ \Phi_t(x+R) - \Phi_t(x) + s(x+R) \Phi_t(x+R) (1-\Phi_t(x)) \right\} \right] + O\left(\frac{1}{N}\right).
\end{align}
This follows in part from $\E[\bar X_\phi] = \phi \mu$ and $\E[\tilde X_\phi] = (1-\phi)\mu$.
We can see two things from this:
First, since this is a first-order differential operator, the limiting stochastic process
is deterministic (check by applying to $f(\phi) = \phi(x)^2$ to find the variance).
Second, if we want to rescale space as well to get the usual differential equation, 
we need to choose $\var[R]=\sigma^2$ and $s(x)$ to be of the same, small, order; 
this is another way of seeing that $\sigma/\sqrt{s}$ is the relevant length scale.
More concretely, suppose that the grid size is $\epsilon \to 0$, that $\var[R] = \sigma^2 \epsilon$, and that $s(x)/\epsilon \to \gamma(x)$,
and suppose that $\Phi_t(x)$ is deterministic (and sufficiently differentiable) and let $\xi(t,x) = \Phi_{t/\epsilon}(x)$;
then the previous equation converges to the familiar form:
\begin{align}
  \partial_t \xi(t,x) = \mu \left( \frac{\sigma^2}{2} \partial_x^2 \xi(t,x) + \gamma(x) \xi(t,x) (1-\xi(t,x)) \right) .
\end{align}

>>>>

Migration--selection balance ensures that 
there will be some number of $B$ alleles present in the regions where they are disadvantageous,
but only very few, and rarely, far away from the patch where $B$ is established.
Denote the expected frequency of allele $B$ at distance $r$ from the patch by $q(r)$.
Following \citet{slatkin1973geneflow}, one can write down a differential equation to which $q(r)$ is the solution,
which can then be solved to give $q(r)$ in terms of Jacobi elliptic functions (see Appendix \ref{apx:elliptic_integrals}).
Asymptotics of these implies that for large $r$ the mean occupation frequency decays exponentially, 
\begin{align} \label{eqn:eqfreq}
    q(r) \approx C \exp( - r \sqrt{2 s_m} / \sigma) \qquad \text{for large $r$},
\end{align}
where $d$ is the dimension ($d=1$ or $2$), and $C$ is a constant depending on the geographic shape of the populations and the selection coefficients. 
In Appendix \ref{apx:patchy_sims} we argue that $C=1/2$ is about right for most parameter values,
and also that these asymptotics are correct in both one and two dimensions
(the differential equation is only analytically solvable in one dimension).

@@

The arguments depends on the following decomposition of the genealogy of $B$ alleles.
First, fix some intermediate distance $r_0$ from the established patch,
and classify all $B$ alleles present at any time outside $r_0$ into families
if they share a common ancestor living outside $r_0$.
If $r_0$ is large enough that there are not too many families
and interactions between family members are weak,
then these families can be approximated by a collection of independent spatial branching processes.
(This can be made formal in the limit of large population density.)
We can extend these families by 
including any individual outside the original patch (but perhaps inside $r_0$) 
that has a parent outside $r_0$,
so that these branching processes are rooted at $r_0$, and are killed upon reaching the original patch.
This decomposition is illustrated in~\ref{fig:sim_snapshots}, and later in figure~\ref{fig:branching_decomp}.
In terms of these families, the equilibrium frequency above is then
\begin{align}
    \label{eqn:gestalt_q}
    \begin{split}
        q(r) &= \text{ ( outflux of families ) } \times \text{ ( mean occupation density of a family at $r$ ) } ,
\end{split}
\end{align}
and the quantity we wish to compute,
the effective rate of at which \emph{families of} migrant $B$ alleles establish in a new patch at distance $r$, 
is
\begin{align}
    \label{eqn:gestalt_migrate}
    \begin{split}
        \migrate(r) &= \text{ ( outflux of families ) } \times \text{ ( probability a family establishes in patch at $r$ ) }
    \end{split}
\end{align}
All terms except the ``outflux of families'' are simple calculations with spatial branching processes;
and we can solve for this outflux using the established form for \eqref{eqn:gestalt_q}.
Here $\migrate$ reflects the mean number of establishing migrant families per unit time,
rather than individuals,
because family establishment events may include more than one individual,
so individual establishment events are clustered in time,
and using their overall density to obtain a typical waiting time until the next event
will underestimate that time.

@@

\begin{figure}[ht!!]
  \begin{center}
    \includegraphics{branching-concept}
  \end{center}
\caption{Cartoon of the decomposition described in the text.
Migrants families are emerging from the already-established patch (grey circle);
each is composed of a ``trunk'' lineage (in red), whose spatial motion we follow,
embellished by transient branches.
For the depicted long-lived family, these are further distinguished by those still alive at time $t$ (in black)
and those that did not survive until $t$ (in grey).
Figure~\ref{fig:sim_snapshots} has a depiction of such a family, from a simulation.
\label{fig:branching_decomp}
}
\end{figure}

@@

\begin{figure}[ht!]
    \begin{center}
        % made in patchy-sim-writeup-plots.R
        \includegraphics{example-equilibrium}
    \end{center}
    \label{fig:example_equil}
    \caption{
    Snapshots of a simulation of allelic frequency dynamics around a patch where the allele is beneficial:
    the focal allele is benefical in the middle region and deleterious outside.
    The dynamics were simulated for 10000 generations across demes with 1000 individuals in each
    and random dispersal to nearby demes with a mean dispersal distance of $\sigma \approx 2$ demes.
    Colored lines show allele frequencies across space in 50 uniformly spaced generations;
    the solid line shows the mean across all generations.
    Below is the same plot, except on a log scale.
    }
\end{figure}

@@

One route to the ``equlibrium frequency'' of the allele outside the range where it is advantageous is as follows;
see \citet{slatkin1973geneflow} or \citet{barton1987establishment,pollak1966survival} for other arguments in this case, 
or \citet{etheridge} or \citet{dawson} for general theory used below.
Suppose that the population is composed of a finite number of small demes of equal size $N$ arranged in a regular grid,
and that selection (for or against) the allele is given by the function $s(x)$, with $x$ denoting the spatial location.
Then each individual at location $x$ reproduces after a random lifetime,
dying and giving birth in the same deme to a random number of offspring with distribution given by $X$;
who all migrate to a new location chosen as $x+R$
and replace randomly chosen individuals there.
If $x+R$ is outside of the range, then they perish.
Each individual's reproduction time is exponentially distributed, 
either with rate 1 if it carries the original allele, or with rate $1+s(x)$ if it carries the mutant allele.
Suppose that $X$ has mean $\mu$; the variance of $X$ will not enter into the formula.
Also suppose that $R$ has mean zero and variance $\sigma^2$;
if we are in more than one dimension, we mean that the components of the dispersal distance are uncorrelated
and each have variance $\sigma^2$.
We also assume that the distribution of $X$ makes the resulting Markov chain irreducible and aperiodic.

Let $\Phi^N_t(x)$ be the number of mutant alleles present at location $x$ at time $t$,
and denote by $\delta_x$ a single unit at location $x$, so that e.g.~$\Phi^N_t + \delta_x/N$
is the configuration after a mutant allele has been added to location $x$.
For $0\le \phi \le 1$, we also denote by $\bar X_\phi$ the number of mutant alleles added if $X$ new offspring carrying mutant alleles
replace randomly chosen individuals in a deme where the mutant allele is at frequncy $\phi$ (i.e.~hypergeometric with parameters $(X,\phi)$);
similarly, $\tilde X_\phi$ is the number lost if the new offspring do not carry the allele (i.e.~hypergeometric with parameters $(X,1-\phi)$).
(We like to think of $\Phi^N_t$ as a measure, but it does not hurt to think of $\Phi^N$ as a vector;
we aren't providing the rigorous justification here.)
Then we know that for any sufficiently nice function $f(\phi)$ that
\begin{align} \label{eqn:discrete_generator}
  \begin{split} \frac{\partial}{\partial t} \E\left[ f(\Phi^N_t) \right] 
  &= N \sum_x \left\{ \E\left[ (1+s(x+R)) \Phi^N_t(x+R) \left( f\left(\Phi^N_t + \frac{\bar X_{\Phi_t(x+R)}}{N}\delta_{x}\right) - f(\Phi^N_t) \right) \right] \right. \\
     & \qquad  \qquad \left. {} + \E\left[ \left(1-\Phi^N_t(x+R)\right) \left( f\left(\Phi^N_t - \frac{\tilde X_{\Phi_t(x+R)}}{N}\delta_{x}\right) - f(\Phi^N_t) \right) \right] \right\}  \end{split} \\
     &= \mu \sum_x \E\left[ \left(\partial_{\phi(x)} f(\Phi_t) \right) \left\{ \Phi_t(x+R) - \Phi_t(x) + s(x+R) \Phi_t(x+R) (1-\Phi_t(x)) \right\} \right] + O\left(\frac{1}{N}\right).
\end{align}
This follows in part from $\E[\bar X_\phi] = \phi \mu$ and $\E[\tilde X_\phi] = (1-\phi)\mu$.
We can see two things from this:
First, since this is a first-order differential operator, the limiting stochastic process
is deterministic (check by applying to $f(\phi) = \phi(x)^2$ to find the variance).
Second, if we want to rescale space as well to get the usual differential equation, 
we need to choose $\var[R]=\sigma^2$ and $s(x)$ to be of the same, small, order; 
this is another way of seeing that $\sigma/\sqrt{s}$ is the relevant length scale.
More concretely, suppose that the grid size is $\epsilon \to 0$, that $\var[R] = \sigma^2 \epsilon$, and that $s(x)/\epsilon \to \gamma(x)$,
and suppose that $\Phi_t(x)$ is deterministic (and sufficiently differentiable) and let $\xi(t,x) = \Phi_{t/\epsilon}(x)$;
then the previous equation converges to the familiar form:
\begin{align}
  \label{eqn:derived_fkpp}
  \partial_t \xi(t,x) = \mu \left( \frac{\sigma^2}{2} \sum_{k=1}^d \partial_{x_k}^2 \xi(t,x) + \gamma(x) \xi(t,x) (1-\xi(t,x)) \right) .
\end{align}

In one dimension, we are done (see section~\ref{apx:elliptic_integrals});
in higher dimensions, we are more interested in the mean frequency at a given distance $r$ from a patch.
If we take a radially symmetric patch centered at the origin (so $\gamma$ only depends on $r$), 
and let $\xi(t,r)$ denote the mean occupation frequency at distance $r$,
then the polar form of the Laplacian in $d$ dimensions gives us that \eqref{eqn:derived_fkpp} is
\begin{align}
  \label{eqn:radial_fkpp}
  \partial_t \xi(t,r) = \mu \left( \frac{\sigma^2}{2} \partial_{r}^2 \xi(t,r) + \sigma^2\frac{d-1}{2r} \partial_r \xi(t,r) + \gamma(r) \xi(t,r) (1-\xi(t,r)) \right) .
\end{align}
However, the second term goes to zero as $r \to \infty$,
leading to the same asymptotic behavior (CITE SOMETHING).
